Exploring the ages of rotating stars using galactic dynamics: a novel approach to calibrating gyrochronology

### 2019-10-15 paper/abstract.tex
many paper edits

The rotational evolution of cool dwarfs is poorly constrained after $\sim$2-3
billion years due to a lack of precise ages and rotation periods for old
main-sequence stars.
% However, recent evidence suggests that the relationship between rotation
% period and color/temperature/mass does not remain constant over large
% timescales, contradicting a long-held assumption that underlies many
% gyrochronology relations.
In this work, we use the velocities of low-mass \kepler\ dwarfs to reveal
their rotational evolution and demonstrate that kinematics have potential to
be a useful tool for calibrating gyrochronology.
At ages less than $\sim$1 Gyr we find that a gyrochronology model, calibrated
to the Praesepe cluster, accurately predicts the relative ages of stars.
At these young ages, lower-mass stars lose angular momentum more quickly than
higher mass stars and therefore rotate more slowly at the same age.
However, at old ages, late G and early K dwarfs rotate {\it faster} than late
K dwarfs of the same age.
This implies that low-mass stars lose angular momentum more slowly than
higher-mass stars after a certain age or Rossby number.
Our results, based on field star rotation periods, align with recent findings
from the rotation periods of stars in middle-aged open clusters and
theoretical models that vary the rate of surface-to-core angular momentum
transport.
Finally, we find no evidence for mass-dependent heating in a sample of K and M
dwarfs in the \kepler\ field.

>>>>

The rotational evolution of cool dwarfs is poorly constrained after $\sim$2-3
billion years due to a lack of precise ages and rotation periods for old
main-sequence stars.
% However, recent evidence suggests that the relationship between rotation
% period and color/temperature/mass does not remain constant over large
% timescales, contradicting a long-held assumption that underlies many
% gyrochronology relations.
In this work, we use the velocities of low-mass \kepler\ dwarfs to reveal
their rotational evolution and demonstrate that kinematics have potential to
be a useful tool for calibrating gyrochronology.
At ages less than $\sim$1 Gyr we find that a gyrochronology model, calibrated
to the Praesepe cluster, accurately predicts the relative ages of stars.
At these young ages, lower-mass stars spin more slowly than higher-mass stars
because their stronger magnetic fields lead to more efficient angular momentum
loss.
However, at old ages we find that late G and early K dwarfs rotate {\it
faster} than late K dwarfs of the same age.
% These oThis may be a feature of angular momentum redistribution in the stellar
% interior.
% This implies that low-mass stars lose angular momentum more slowly than
% higher-mass stars after a certain age or Rossby number.
These results, based on field star rotation periods, align with recent
findings from the rotation periods of stars in middle-aged open clusters and
theoretical models that vary the rate of surface-to-core angular momentum
transport as a function of time and mass.
Finally, we find no evidence for mass-dependent heating in a sample of K and M
dwarfs in the \kepler\ field.

### 2019-10-15 paper/conclusion.tex
many paper edits

We examined the gyrochronology relations using the velocity dispersions of
field stars with measured rotation periods.
We found that old groups of cool dwarfs selected to be coeval using the
\citet{angus2019} gyrochronology relation do {\it not} have the same velocity
dispersion across all temperatures.
This implies that the \citet{angus2019} relation, which is based on the
period-color relation of the 650 Myr Praesepe cluster, does not correctly
describe the period-age-color/\teff\ relation for old stars.
It appears that the period-color/\teff\ relation changes shape over time, \ie\
the period-color/\teff\ relation and the period-age relation are
non-separable.
At ages younger than $\sim$1 Gyr, rotation period is anti-correlated with
\teff: cooler stars spin more slowly than hotter stars of the same age.
However, at old ages the opposite is true: cooler stars spin more rapidly than
hotter stars of the same age.
The period-color/\teff\ relation appears to start flattening out after $\sim$1
Gyr (see figure \ref{fig:age_cut}), potentially around the same age, or just
older than the period gap which is located at a gyrochronal age of around
$\sim$1.1 Gyr (see figure \ref{fig:dispersion_period_teff}).
This gap may represent a significant transitional epoch in the magnetic
behavior of stars.
Finally, when velocity dispersion is interpreted as an age proxy, it appears
that the oldest stars in the \mct\ catalog are cooler than 4500 K, which
suggests that lower-mass stars remain active for longer and their rotation
periods can be measured at older ages.
We outlined a number of scenarios which could provide alternative
explainations for these observations, including incorrect dust corrections for
the lowest-mass stars and an excess of companions increasing the velocity
dispersion for these stars.

If the period-color/\teff\ relation does invert at old ages as our results
suggest, this would be a paradigm shift for gyrochronology.
Stellar spin-down rate is thought to be directly tied to magnetic field
strength, and the deeper convection zones of cooler stars generate stronger
magnetic fields which {\it should} lead to more efficient angular momentum
loss.
However, the micro- and macro-physics of stellar structure and evolution and
magnetic dynamo models are extremely complicated and a lot is still unknown
about the magnetic behavior of stars.
Observations like these can provide useful constraints for physical models,
and may help to reveal new physical processes at work in stars like our own
Sun and other planet hosts.

>>>>

We examined the rotational evolution of \kepler\ field stars using their
velocity dispersions as a proxy for age.
We found that old groups of cool dwarfs selected to be coeval using the
\citet{angus2019} gyrochronology relation do {\it not} have the same velocity
dispersion across all temperatures.
This implies that the \citet{angus2019} relation, which is based on the
period-color relation of the 650 Myr Praesepe cluster, does not correctly
describe the period-age-color/\teff\ relation for old stars.
It appears that the period-color/\teff\ relation changes shape over time in a
way that qualitatively agrees with theoretical that which include a
mass-dependent core-envelope angular momentum transport \citep{spada2019}.
At young ages, rotation period is anti-correlated with \teff: cooler stars
spin more slowly than hotter stars of the same age.
However, at intermediate ages the relation flattens out and K dwarfs rotate at
the same rate, regardless of mass.
At old ages, it seems that cooler K dwarfs spin more rapidly than hotter K
dwarfs of the same age.
We speculate that the rotation period gap \citep{mcquillan2014} may separate
a young regime where stellar rotation periods decrease with increasing mass
from an old regime where periods increase with increasing mass, however more
data are needed to provide a conclusive result.
% The period-color/\teff\ relation seems to start flattening out after $\sim$1
% Gyr (see figure \ref{fig:age_cut}), potentially around the same age, or just
% older than the period gap which is located at a gyrochronal age of around
% $\sim$1.1 Gyr (see figure \ref{fig:dispersion_period_teff}).
Finally, when velocity dispersion is interpreted as an age proxy, it appears
that the oldest stars in the \mct\ catalog are cooler than 4500 K, which
suggests that lower-mass stars remain active for longer, allowing their
rotation periods to be measured at older ages.
% We outlined a number of scenarios which could provide alternative
% explanations for these observations, including incorrect dust corrections for
% the lowest-mass stars and an excess of companions increasing the velocity
% dispersion for these stars.

% If the period-color/\teff\ relation does invert at old ages as our results
% suggest, this would be a paradigm shift for gyrochronology.
% Stellar spin-down rate is thought to be directly tied to magnetic field
% strength, and the deeper convection zones of cooler stars generate stronger
% magnetic fields which {\it should} lead to more efficient angular momentum
% loss.
% However, the micro- and macro-physics of stellar structure and evolution and
% magnetic dynamo models are extremely complicated and a lot is still unknown
% about the magnetic behavior of stars.
% Observations like these can provide useful constraints for physical models,
% and may help to reveal new physical processes at work in stars like our own
% Sun and other planet hosts.

### 2019-10-15 paper/discussion.tex
modifying paper text and experimenting with various cuts (the overall results are the same)
many paper edits

\subsection{The period gap}
\label{sec:period_gap}

The origin of the rotation period gap, first identified
by \citet{mcquillan2013} and visible in figures \ref{fig:age_cut} and
\ref{fig:dispersion_period_teff} still remains a mystery.
This gap can be seen as an under-density of points between the 0.7-1.0 and
1.0-1.5 Gyr age ranges in figure \ref{fig:age_cut} and roughly follows a line
of constant gyrochronal age of around 1.1 Gyr \citep[according to the
gyrochronology relation of][]{angus2019}, as shown in figure
\ref{fig:dispersion_period_teff}.
Several explanations for the gap's origin have been proposed, including a
discontinuous star formation history \citep{mcquillan2013, davenport2017,
davenport2018}, a rapid change in magnetic field structure
\citep{reinhold2019}, and erroneous rotation period measurements that are
incorrect by a factor of two \citep{koen2018}.
The latter explanation can be ruled out because stars below the gap have
smaller velocity dispersions than the stars above the gap, indicating that
they are kinematically younger \citep{mcquillan2013, davenport2018}, as
evident in figures \ref{fig:age_cut} and \ref{fig:dispersion_period_teff}.
For stars below the gap, in the 0.7-1.0 Gyr age range shown in figure
\ref{fig:age_cut}, velocity dispersion is relatively constant as a function of
temperature, however above the gap, in the 1.0-1.5 Gyr age range and older,
velocity dispersion increases with \teff.
The coolest stars in the 1.0-1.5 Gyr age range have the same velocity
dispersion as the hottest stars in the age range above which indicates that
the period-\teff\ relations are {\it flat} at these rotation periods.
This is also visible in figure \ref{fig:dispersion_period_teff}.
Below the gap, velocity dispersion within a given period range appears to {\it
decrease} with decreasing temperature.
The opposite appears to be true above the gap.
It could be that the gap is positioned at a significant Rossby number/age at
which stellar magnetic dynamos go through a transition.
Perhaps before the age of 1.1 Gyr, or at Rossby numbers less than ???,
magnetic braking is more efficient for stars with deeper convection zones.
Once stars reach this critical age or Rossby number their magnetic fields
undergo some radical transition, which produces the gap in the rotation
period-\teff\ plane.
After this transition, magnetic braking efficiency no longer increases with
decreasing mass.
Of course, it may be a coincidence that the gyrochronology relations seem to
only flatten off above the period gap and we lack a sufficient quantity of
data to do more than speculate here.
New rotation periods from the \ktwo\ and \tess\ missions may be able to
validate or rule out this hypothesis in the future.

In the $\sim$ 1.1 Gyr NGC 6811 cluster, the rotation periods of mid-K dwarfs
are faster than expected; their rotational evolution appears to have stalled,
and they have similar rotation periods to the 650 Myr Praesepe cluster
\citep{curtis2019}.
The rotation periods of the K dwarfs in this cluster are plotted in figure
\ref{fig:dispersion_period_teff}.
Although NGC 6811's G dwarfs fall on the 1.1 Gyr gyrochronology model, the K
dwarfs lie only a little above the 0.65 Gyr gyrochronology model.
NGC 6811 straddles the rotation period gap: its G dwarfs lie above it and its
K dwarfs lie below it.
This cluster may be the `missing link' that connects two epoch of stellar
spin-down.
However, figure \ref{fig:age_cut} indicates that the Praesepe gyrochronology
model is, on average, a good model for stars younger than $\sim$1 Gyr. This
model has a very different shape to the NGC 6811 cluster.
This would suggest that stars in the field do follow the period-color/\teff\
relation of Praesepe, at least up to around 1 Gyr.
We do not find evidence of the K dwarf stalling, which is observed in NGC
6811, in the field.
% , however without more observations of middle-aged open clusters it : an
% early stage where the period-\teff\ relation for cool dwarfs has a negative
% slope and a late stage where it has a positive slope.
% In this case the period gap may delineate the transition between these two
% regimes and is the point at which stellar magnetic dynamos likely undergo a
% dramatic structural shift at an age of $\sim$ 1.1 Gyr.

>>>>

% \subsection{The period gap}
% \label{sec:period_gap}

The results presented above indicate that stars of spectral type ranging from
late G to late K follow a braking law that changes over time.
These results provide further evidence for `stalled' rotational evolution of K
dwarfs, like that observed in open clusters \citep{curtis2019} and reproduced
by models that vary angular momentum transport between stellar core and
envelope with time and mass \citep{spada2019}.

Our results indicate that, at young ages, stellar rotation period decreases
with mass.
This is probably because lower-mass stars with deeper convection zones have
stronger magnetic fields, larger Alfven radii and therefore experience greater
angular momentum loss rate.
According to the \citet{spada2019} models, there is minimal transportation of
angular momentum from the surface to the core of the star at these young ages,
so the surface slows down but the core keeps spinning rapidly.
At intermediate ages, rotation periods appear constant with mass, and at late
ages rotation period increases with mass for GK dwarfs.
The explanation for this phenomenon, according to the \citet{spada2019}
models, is that lower-mass stars are still braking more efficiently at these
intermediate and old ages but their cores are more tightly coupled with their
envelopes, so angular momentum resurfaces and prevents the stellar envelopes
from spinning-down rapidly.

The origin of the rotation period gap, first identified by
\citet{mcquillan2013} and visible in figures \ref{fig:age_cut} and
\ref{fig:dispersion_period_teff} still remains a mystery.
This gap can be seen as an under-density of points between the 0.7-1.0 and
1.0-1.5 Gyr age ranges in figure \ref{fig:age_cut} and roughly follows a line
of constant gyrochronal age of around 1.1 Gyr \citep[according to the
gyrochronology relation of][]{angus2019}, as shown in figure
\ref{fig:dispersion_period_teff}.
Several explanations for the gap's origin have been proposed, including a
discontinuous star formation history \citep{mcquillan2013, davenport2017,
davenport2018} and a change in magnetic field structure causing a brief period
where rotational variability is reduced and rotation periods cannot be
measured \citep{reinhold2019}.
Our results indicate there might be a slight difference in rotational
evolution below and above the gap.
Stars below the gap appear to be in the regime where wind-braking is dominant
(rotation period decreases with increasing mass at a given age) and stars
above the gap are in a regime where rotational coupling is dominant (rotation
period increases with increasing mass at a given age).
This can be seen in figures \ref{fig:age_cut} and \ref{fig:vplot}.
For stars below the gap, in the 0.7-1.0 Gyr age range shown in figure
\ref{fig:age_cut}, velocity dispersion is relatively constant as a function of
temperature, however above the gap, in the 1.0-1.5 Gyr age range and older,
velocity dispersion increases with \teff.
In figure \ref{fig:age_cut}, velocity dispersion within a given period range
appears to {\it decrease} with decreasing temperature below the gap.
The opposite appears to be true above the gap.
This suggests that perhaps stars transition from a wind braking regime to a
rotational coupling regime as they pass across the gap.
However, it is unclear how this process could {\it create} the gap.
A gap could be created if stars rapidly lose angular momentum over a short
period of time, leading to a dearth of stars in the gap.
However, a sudden redistribution of angular momentum from the core to the
surface, should cause stellar rotation periods to {\it decrease}, not
increase.
In addition, this theory is not supported by observations of the NGC 6811
cluster whose K dwarf members have already undergone stalled braking below and
before they reach the gap.
The gap may be associated with a wind braking-rotational coupling transition,
however our results only provide weak evidence for this, so we can only
speculate until more data become available.

% The \citet{spada2019} models do predict that stars transition from a regime
% where wind-braking is dominant to a regime where rotational coupling is
% domimant at different times for different masses.
% However, these models are based on observations of NGC 6811, whose K dwarfs
% already show signs of stalling by 1.1 Gyr.
% For stellar masses of 0.7 M$_\odot$, this transition takes place at $\sim$ 1
% Gyr.
% It happens at 500 Myr for 0.9 M$_\odot$ and 5 Gyr for 0.5 M$_\odot$
% \citep[see][for details]{spada2019}.
% We propose a new theory for the origin of the rotation gap as the transition
% between the wind-braking and rotational coupling dominated regimes.
% The gap could be caused by rapidly transitioning across the gap
% as

% These results are extremely tentative and should be interpreted with caution:
% more data are needed to confirm this behavior.

% For stars below the gap, in the 0.7-1.0 Gyr age range shown in figure
% \ref{fig:age_cut}, velocity dispersion is relatively constant as a function of
% temperature, however above the gap, in the 1.0-1.5 Gyr age range and older,
% velocity dispersion increases with \teff.
% The coolest stars in the 1.0-1.5 Gyr age range have the same velocity
% dispersion as the mid-temperature stars in the age range above, and the
% hottest stars in the age range above {\it that}, which indicates that the
% period-\teff\ relations are {\it flat} at rotation periods between $\sim$15-25
% days, for 5500 K $<$ \teff\ $<$ 4000 K.
% Below the gap, velocity dispersion within a given period range appears to {\it
% decrease} with decreasing temperature.
% The opposite appears to be true above the gap.

% The gap may be positioned at a significant Rossby number/age at which stellar
% magnetic dynamos go through a transition.
% Perhaps before the age of 1.1 Gyr, or at Rossby numbers less than ???,
% magnetic braking is more efficient for stars with deeper convection zones.
% Once stars reach this critical age or Rossby number their magnetic fields
% undergo some transition, which produces the gap in the rotation period-\teff\
% plane and after this transition, magnetic braking efficiency no longer
% increases with decreasing mass.
% Of course, it may be a coincidence that the gyrochronology relations seem to
% only flatten off above the period gap and we lack a sufficient quantity of
% data to do more than speculate here.
% New rotation periods from the \ktwo\ and \tess\ missions may be able to
% validate or rule out this hypothesis in the future.

% In the $\sim$ 1.1 Gyr NGC 6811 cluster, the rotation periods of mid-K dwarfs
% are faster than expected; their rotational evolution appears to have stalled,
% and they have similar rotation periods to the 650 Myr Praesepe cluster
% \citep{curtis2019}.
% The rotation periods of the K dwarfs in NGC 6811 are plotted in figure
% \ref{fig:dispersion_period_teff}.
% Although NGC 6811's G dwarfs fall on the 1.1 Gyr gyrochronology model, the K
% dwarfs lie only a little above the 0.65 Gyr gyrochronology model.
% NGC 6811 straddles the rotation period gap: its G dwarfs lie above it and its
% K dwarfs lie below it.
% This cluster may be the `missing link' that connects two epoch of stellar
% spin-down.
% However, figure \ref{fig:age_cut} indicates that the Praesepe gyrochronology
% model is, on average, a {\it good} model for stars younger than $\sim$1 Gyr
% % and this model has a very different shape to the NGC 6811 cluster.
% suggesting that stars in the field follow the period-color/\teff\ relation of
% Praesepe, at least up to around 1 Gyr.
% We do not see strong evidence of K dwarf stalling in the field.
% % , however without more observations of middle-aged open clusters it : an
% % early stage where the period-\teff\ relation for cool dwarfs has a negative
% % slope and a late stage where it has a positive slope.
% % In this case the period gap may delineate the transition between these two
% % regimes and is the point at which stellar magnetic dynamos likely undergo a
% % dramatic structural shift at an age of $\sim$ 1.1 Gyr.

### 2019-10-15 paper/method.tex
adding a bunch of paper files -- not sure why these weren't staged?
modifying paper text and experimenting with various cuts (the overall results are the same)
many paper edits

The {\tt Pyia} \citep{price-whelan_2018} and {\tt astropy} \citep{astropy2013,
astropy2018} {\it Python} packages were used to calculate stellar velocities.
{\tt Pyia} calculates velocity samples from the full \gaia\ uncertainty
covariance matrix via Monte Carlo sampling.
It therefore not only incorporates uncertainties on the \gaia\ positions
parallaxes and proper motions, it also accounts for the {\it covariance}
between these properties.
Stars with negative parallaxes, parallax signal-to-noise ratios less than 10,
stars fainter than 16th magnitude, stars with absolute \vb\ uncertainties
greater than 1 \kms\, and stars with galactic latitudes greater than
15\degrees\ were removed from the sample.
Because \vb\ is only a close approximation to \vz\ at low galactic latitudes,
and because latitude is correlated with stellar mass (lower mass stars are
older and tend to be preferentially located at high $b$, although this trends
is greatly reduced for stars cooler than 5000 K), we found that excluding
stars with $b > 15$\degrees\ reduced the number of \vb\ outliers at low
stellar mass.
15\degrees\ was found to be an adequate compromise between maintaining a close
relationship between \vb\ and \vz\ and retaining a sample large enough to
provide meaningful results.
Reducing this cut to 10\degrees\ resulted in a sample size too small to reveal
trends in the data, and intermediate cuts did not significantly change the
results.
% We also found that the anti-correlation between stellar mass and $b$ is
% negligable for stars cooler than 5000 K.
% For this reason, and because the rotational evolution of old, G stars does not
% follow a simple Skumanich-like spin-down law \citep{vansaders2016}, we
% restricted our sample to temperatures between \tmin\ and 5000 K, \ie\ K
% dwarfs.

\subsection{Establishing \vb\ as an age proxy}
\label{sec:mass-dependent-heating}

If lower-mass stars experience greater velocity changes when gravitationally
perturbed and are dynamically heated more efficiently than higher-mass stars,
velocity dispersion would be a function of {\it both} age and mass and cannot
be straightforwardly interpreted as an age proxy.
So, in order to establish whether \sigmavb\ can be used as an age proxy, we
searched for signs of mass-dependent heating within the \kepler\ field.

Mass-dependent dynamical heating has not been unambiguously observed in the
galactic disk because of the strong anti-correlation between stellar mass and
stellar age.
Less massive stars do indeed have larger velocity dispersions, however they
are also older on average.
This mass-age degeneracy is highly reduced in M dwarfs because their
main-sequence lifetimes are longer than the age of the Universe, however no
evidence for mass-dependent heating has been found in M dwarfs
\citep{faherty2009}.

To investigate whether mass-dependent heating could be acting on the \kepler\
sample, we selected late K and M dwarfs observed by both \kepler\ and \gaia,
whose MS lifetimes exceed around 11 Gyrs and are therefore representative of
the initial mass function.
We could not perform this analysis on the \mct\ sample, because only stars
with {\it detectable} rotation periods appear there.
Since lower-mass stars stay active for longer it is likely that there are more
old stars (with larger velocity dispersions) at low masses in this sample.
We selected all \kepler\ targets with dereddened \gaia\ \gcolor\ colors
greater than 1.2 (corresponding to an effective temperature $\lesssim$
4800 K) and absolute \gaia\ $G$-band magnitudes $<$ 4.
We also eliminated visual binaries by removing stars above a 6th order
polynomial, fit to the MS on the \gaia\ CMD.
We then applied the quality cuts described in section \ref{sec:the_data}.
To search for evidence of mass-dependent heating we calculated the (\vb)
velocity dispersion of stars in effective temperature bins.
Sigma clipping was performed at the 3$\sigma$ level to remove high velocity
outliers before calculating the standard deviation of stars in each bin.

Figure \ref{fig:vb_vs_teff} shows velocity and velocity dispersion
as a function of effective temperature (calculated by transforming dereddened
\gaia\ colors using equation \ref{eq:curtis}).
Velocity dispersion very slightly {\it decreases} with decreasing temperature,
the opposite of the trend expected for mass-dependent heating, however the
slope is only inconsistent with zero at the 1.3 $\sigma$ level.
This trend may be due to a selection bias: cooler stars are fainter and
therefore typically closer, with smaller heights above the galactic plane and
smaller velocities.
The essential point is that we do not find evidence for mass-dependent heating
acting on stars in the \kepler\ field, meaning \sigmavb\ {\it can} be used as
an age proxy.
We performed the same analysis on the 537 stars in this sample with RVs using
{\it vertical} velocity (\vz) and found the slope of the velocity
dispersion-temperature relation was consistent with zero.

Since we used \vb, not \vz\ in our analysis, it is possible that an
anti-correlation between \teff\ and galactic latitude, $b$, could influence
our results.
Stars at higher latitudes have additional velocity components in the ${\bf x}$
and ${\bf y}$ directions, which could increase \vb\ but not \vz.
Again however, since the relationship between \sigmavb\ and \teff\ is
positively, not negatively correlated for cool stars in the \kepler\ field,
this effect is probably too small to influence our results.

>>>>

The {\tt Pyia} \citep{price-whelan_2018} and {\tt astropy} \citep{astropy2013,
astropy2018} {\it Python} packages were used to calculate stellar velocities.
{\tt Pyia} calculates velocity samples from the full \gaia\ uncertainty
covariance matrix via Monte Carlo sampling.
It therefore not only incorporates uncertainties on the \gaia\ positions
parallaxes and proper motions, it also accounts for the {\it covariance}
between these properties.
Stars with negative parallaxes, parallax signal-to-noise ratios less than 10,
stars fainter than 16th magnitude, stars with absolute \vb\ uncertainties
greater than 1 \kms\, and stars with galactic latitudes greater than
15\degrees\ (justification provided below) were removed from the sample.
% Because \vb\ is only a close approximation to \vz\ at low galactic latitudes,
% and because latitude is correlated with stellar mass (lower mass stars are
% older and tend to be preferentially located at high $b$, although this trends
% is greatly reduced for stars cooler than 5000 K), we found that excluding
% stars with $b > 15$\degrees\ reduced the number of \vb\ outliers at low
% stellar mass.
% 15\degrees\ was found to be an adequate compromise between maintaining a close
% relationship between \vb\ and \vz\ and retaining a sample large enough to
% provide meaningful results.
% Reducing this cut to 10\degrees\ resulted in a sample size too small to reveal
% trends in the data, and intermediate cuts did not significantly change the
% results.
% We also found that the anti-correlation between stellar mass and $b$ is
% negligable for stars cooler than 5000 K.
% For this reason, and because the rotational evolution of old, G stars does not
% follow a simple Skumanich-like spin-down law \citep{vansaders2016}, we
% restricted our sample to temperatures between \tmin\ and 5000 K, \ie\ K
% dwarfs.

\subsection{Validating \vb\ dispersion as an age proxy}
\label{sec:mass-dependent-heating}

There are two main reasons why \vb\ velocity dispersion may not be a good age
proxy.
Firstly, mass-dependent heating may act on the sample, meaning that velocity
dispersion depends on both age and mass, so cannot be interpreted as a simple
age proxy.
Secondly, since stars in the \kepler\ field have a range of galactic
latitudes, using \vb\ as a stand-in for \vz\ may not be equally valid for all
stars, introducing a velocity bias for high latitude stars (which are more
likely to be cooler and older).
In this section we demonstrate that neither of these problems seem to be a
significant issue for our data.

If lower-mass stars experience greater velocity changes when gravitationally
perturbed and are dynamically heated more efficiently than higher-mass stars,
velocity dispersion would be a function of {\it both} age and mass and cannot
be straightforwardly interpreted as an age proxy.
So, in order to establish whether \sigmavb\ can be used as an age proxy, we
searched for signs of mass-dependent heating within the \kepler\ field.
Mass-dependent dynamical heating has not been unambiguously observed in the
galactic disk because of the strong anti-correlation between stellar mass and
stellar age.
Less massive stars do indeed have larger velocity dispersions, however they
are also older on average.
This mass-age degeneracy is highly reduced in M dwarfs because their
main-sequence lifetimes are longer than the age of the Universe, however no
evidence for mass-dependent heating has been found in M dwarfs
\citep[\eg][]{faherty2009, newton2016}.

To investigate whether mass-dependent heating could be acting on the \kepler\
sample, we selected late K and M dwarfs observed by both \kepler\ and \gaia,
whose MS lifetimes exceed around 11 Gyrs and are therefore representative of
the initial mass function.
We could not perform this analysis on the \mct\ sample, because only stars
with {\it detectable} rotation periods are included in it.
Since lower-mass stars stay active for longer it is likely that the oldest
stars in this sample have low masses.
We selected all \kepler\ targets with dereddened \gaia\ \gcolor\ colors
greater than 1.2 (corresponding to an effective temperature $\lesssim$
4800 K) and absolute \gaia\ $G$-band magnitudes $<$ 4.
We also eliminated visual binaries by removing stars above a 6th order
polynomial, fit to the MS on the \gaia\ CMD.
We then applied the quality cuts described above in section
\ref{sec:the_data}.
To search for evidence of mass-dependent heating we calculated the (\vb)
velocity dispersion of stars in effective temperature bins.
Sigma clipping was performed at the 3$\sigma$ level to remove high velocity
outliers before calculating the standard deviation of stars in each bin.
These high velocity outliers may be very old late K and M dwarfs, or they
result from using \vb\ instead of \vz, which introduces additional velocity
scatter.

Figure \ref{fig:vb_vs_teff} shows velocity and velocity dispersion
as a function of effective temperature (calculated by transforming dereddened
\gaia\ colors using equation \ref{eq:curtis}).
Velocity dispersion very slightly {\it decreases} with decreasing temperature,
the opposite of the trend expected for mass-dependent heating, however the
slope is only inconsistent with zero at the 1.3 $\sigma$ level.
This trend may be due to a selection bias: cooler stars are fainter and
therefore typically closer, with smaller heights above the galactic plane and
smaller velocities.
The essential point is that we do not find evidence for mass-dependent heating
acting on stars in the \kepler\ field, meaning \sigmavb\ {\it can} be used as
an age proxy.
We also examined the {\it vertical} velocities of the 537 stars in this sample
with RV measurements.
Again, we found no evidence for mass-dependent heating: the slope of the
velocity dispersion-temperature relation was consistent with zero.

At a galactic latitude of zero, $v_b=v_z$, however for increasing values of
$b$, this equivalence becomes an approximation that grows noisier with $b$.
To test the validity of the \vb$\sim$\vz\ approximation over a range of
latitudes we downloaded stellar data from the \Gaia\ Universe Model Snapshot
(GUMS) simulation -- a simulated \Gaia\ catalog \citep{robin2012}.
We downloaded stars from four pointings in the \kepler\ field with galactic
latitudes of around 5\degrees, 10\degrees, 15\degrees, and 20\degrees, out to
a limiting magnitude of 16 dex.
We found that \vb\ is drawn from a heavy-tailed distribution, centered on \vz,
with a standard deviation (calculated as 1.5 $\times$ the median absolute
deviation) that increased with increasing $b$.
The standard deviation of \vz-\vb\ was around 3kms$^{-1}$ at $b \sim 5^\circ
$, 4kms${^-1}$ at 10$^\circ$, 6kms${^-1}$ at 15$^\circ$, and 9kms${^-1}$ at
20$^\circ$.
Since we are concerned with velocity {\it dispersions}, rather than velocities
themselves, we also compared the \vb\ and \vz\ velocity dispersions for stars
downloaded from the GUMS simulation.
For stars at galactic latitudes of 15\degrees\ or less, \sigmavb\ was
consistent with $\sigma_{v{\bf z}}$, within uncertainties, however, at higher
latitudes the two quantities became significantly different.
For this reason we proceeded by only including stars with galactic latitudes
less than 15\degrees\ in our analysis.

% Since we use \vb, not \vz\ in our analysis, it is possible that the sample
% selection function could influence our results.
% For example, higher-mass stars tend to be younger and reside closer to the
% galactic disk mid-plane (low galactic latitudes).
% Since \vb\ is only similar to \vz\ at low latitudes, this could mean that
% lower-mass stars typically have greater velocity dispersions due to .
% \teff\ and galactic latitude, $b$,
% could result in
% Stars at higher latitudes have additional velocity components in the ${\bf x}$
% and ${\bf y}$ directions, which could increase \vb\ but not \vz.
% Again however, since the relationship between \sigmavb\ and \teff\ is
% positively, not negatively correlated for cool stars in the \kepler\ field,
% this effect is probably too small to influence our results.
